---
title: "Example 1: Cloud-free composite images"
execute: 
  eval: true # local
---

```{r}
#| label: config
#| include: false
if (file.exists("config.R")) {
  source("config.R")
}
```

```{r}
#| include: false
gdalcubes_options(default_chunksize = c(1,128,128))
```

*Please notice that code chunks in this document are meant to be executed on an Amazon Web Services (AWS) machine in region `us-west-2` (Oregon). Examples have been selected to yield computation times acceptable for a live demonstration. Please feel free to apply on larger areas and/or using a higher spatial resolution.*

## Introduction

In this example, we will create a cloud-free RGB composite image for the [Córdoba Province, Argentina](https://en.wikipedia.org/wiki/C%C3%B3rdoba_Province,_Argentina) from a collection of Seninel-2 images.
We will read imagery from the [open Sentinel-2 COG catalog on AWS](https://registry.opendata.aws/sentinel-2-l2a-cogs/) and use the available
[Earth Search STAC-API service](https://earth-search.aws.element84.com/v0) to search for images intersecting with our region of interest. 


```{r}
#| include: false
library(tmap)
tmap_options(basemaps = c(OSM = "OpenStreetMap", Imagery = "Esri.WorldImagery")) # TODO: put in config
tmap_mode("view")
```


## 1. Define area of interest 

Our area of interest is provided as a GeoPackage file (data/cordoba_region.gpkg), containing a single polygon. 
We can use the `sf` [@sf] package to read the file afterwards use the `tmap` [@tmap] package to create a simple map.

```{r}
library(sf)
cordoba_shape = read_sf("data/cordoba_region.gpkg")

library(tmap)
tm_shape(st_geometry(cordoba_shape)) +  
  tm_polygons(alpha = 0.2, col = "red")
```



## 2. Query available images from STAC

To find images that intersect with our region and time of interest, we can extract the bounding box of our 
polygons with `st_bbox()`. The STAC request, however, expects WGS84 coordinates and we therefore transform (`st_transform`) the polygon and derive the bounding box again. 


```{r}
bbox = st_bbox(cordoba_shape) 
cordoba_shape |>
  st_transform("EPSG:4326") |>
  st_bbox() -> bbox_wgs84
bbox_wgs84
```

Next, we can load the `rstac` package, connect to the Earth Search service and request images that intersect with
the derived bounding box and the provided time range (March 2022).

```{r}
library(rstac)
s = stac("https://earth-search.aws.element84.com/v0")
items = s |>
  stac_search(collections = "sentinel-s2-l2a-cogs",
              bbox = c(bbox_wgs84["xmin"],bbox_wgs84["ymin"],
                       bbox_wgs84["xmax"],bbox_wgs84["ymax"]), 
              datetime = "2022-03-01/2022-03-31") |>
  post_request() |> items_fetch(progress = FALSE)
items
```

The result contains 385 images as a list `items$features`. Each item (image) is in turn a list with properties, 
assets (links to S3 objects / files), and other metadata. Below, we show some example fields of a single item.

```{r}
items$features[[20]]$properties
items$features[[20]]$assets$B08
```


## 3. Create a gdalcubes image collection from STAC result

To create a data cube from the images contained in the STAC response, we can now use the `gdalcubes` package.
First, we need to convert the STAC item list to a gdalcubes _image collection object_, which indexes available images (in a single-file database). Image collection objects do not contain any pixel data and hence are very small. Below, we use the `stac_image_collection()` function that receives a list of STAC items as input and returns a gdalcubes image collection as output. We explicitly provide names of assets to make sure that the "SCL" band is included and at the same time apply a filter function on images to ignore cloudy images.


```{r}
library(gdalcubes)
assets = c("B01","B02","B03","B04","B05","B06", "B07","B08","B8A","B09","B11","SCL")
s2_collection = stac_image_collection(items$features, asset_names = assets, 
                                      property_filter = function(x) {x[["eo:cloud_cover"]] < 20})
s2_collection
```



## 4. Create a (virtual) data cube

We can now define our target data cube, including the coordinate reference system, the pixel sizes, the spatiotemporal extent as well as methods used for spatial resampling and temporal aggregation. 


```{r}
v= cube_view(srs="EPSG:3857",  dx=500, dy=500, dt="P1D", 
             aggregation="median", resampling = "average",
             extent=list(t0 = "2022-03-01", t1 = "2022-03-31",
                         left=bbox["xmin"], right=bbox["xmax"],
                         top=bbox["ymax"], bottom=bbox["ymin"]))
v
```

To ignore cloud and cloud shadow pixels during construction of the data cube, we define a mask using the [SCL band](https://sentinels.copernicus.eu/web/sentinel/technical-guides/sentinel-2-msi/level-2a/algorithm) and afterwards combine the collection, the data cube view, and the mask to build a data cube.  

```{r}
S2.mask = image_mask("SCL", values = c(3,8,9))
S2_cube = raster_cube(s2_collection, v, S2.mask)
S2_cube
```

Notice that the result is a _virtual_ data cube (or a _proxy_ object) that still does not contain any pixel data and consumes any processing / memory resources. Instead, the object simply knows _how_ to create the cube when needed (e.g. when you call `plot()`, `write_tif()`, `write_ncdf()`, or similar). This concept is sometimes referred to as _lazy evaluation_.


## 5. Process data cube and plot result

Given the data cube, we can apply built-in data cube operations. Notice that calling these operations still will
not start any expensive computations / data reading and the returned object is still a _virtual_ data cube. 
To derive a simple cloud-free mosaic image of our study region, we use the `select_bands()` function to filter by spectral bands, the `reduce_time()` function to calculate median values for all pixel time series and all three bands, and the `filter_geom()` function to cut our polygon from the result. 
Calling `plot()` will eventually start needed computations and finally plot a composite image. 


```{r}
S2_cube |>
  select_bands(c("B02", "B03", "B04")) |>
  reduce_time("median(B02)","median(B03)","median(B04)") |>
  filter_geom(cordoba_shape$geom) |>
  plot(rgb = 3:1, zlim = c(0, 1800))
```

Notice that computations are executed in parallel. Calling `gdalcubes_options(parallel = 8)` can be used to e.g. use up to 8 parallel worker processes.


In addition to the operations used above, the `gdalcubes` package provides the following operations on data cubes.


| Operation           | Description                                                                        | 
|:--------------------|:-----------------------------------------------------------------------------------|    
|`aggregate_space`    | Reduce spatial resolution of a cube by applying a spatial aggregation function. |
|`aggregate_time`     | Aggregate and/or regularize time series. |
|`apply_pixel`        | Apply an arithmetic expression to all data cube pixels.    |
|`crop`               | Extract a rectangular spatial / temporal / spatiotemporal window. |
|`fill_time`          | Fill missing values of a data cube by simple time series interpolation. |
|`filter_geom`        | Filter pixels by a a spatial polygon. |
|`filter_pixel`       | Filter pixels by a logical expressions on band values. |
|`join_bands`         | Combine bands of two identically shaped data cubes. |
|`reduce_space`       | Apply a reducer function to all spatial slices of a data cube.  |
|`reduce_time`        | Apply a reducer function to all pixel time series. |
|`select_bands`       | Select specific bands of a data cube.  |
|`select_time`        | Select irregular time slices of a data cube. |
|`slice_space`        | Select a single time series of a data cube. |
|`slice_time`         | Select a single time slice of a data cube. |
|`window_time`        | Apply a moving window aggregate or convolution kernel to all pixel time series. |

: Built-in data cube operations {#tbl-operations}




## 6. Greenest pixel composite with a user-defined function  

Some operations accept user-defined R functions as arguments. For example, we can write
our own reducer function that computes per pixel NDVI values, and returns RGB values at the day
with maximum NDVI (_greenest_ pixel) as in the following code example.

```{r}
raster_cube(s2_collection, v, S2.mask) |>
  select_bands(c("B02", "B03", "B04", "B08")) |>
  reduce_time(names=c("blue", "green", "red"), FUN = function(x) {
    ndvi = (x["B08", ] - x["B04", ]) / (x["B08", ] + x["B04", ])
    if (all(is.na(ndvi))) {
      return(c(NA,NA,NA))
    }
    i = which.max(ndvi)
    return(x[c("B02", "B03", "B04"), i])
  }) |>
  filter_geom(cordoba_shape$geom) |>
  plot(rgb = 3:1, zlim = c(0, 1800))
```

